---
title: "Assignment 5: Substance Abuse Treatment Mapping"
subtitle: "STUDENT NAME"
author: "Amber Camp"
date: "`r Sys.Date()`"
format: html
editor: visual
---

## Overview

Due: **Tuesday, 10/7 11:59pm**

In this assignment, you will apply analytic and mapping skills on the TEDS-D dataset from SAMHSA.

Refer to the following files for reference, though some of what is asked of you in this assignment will require logic beyond what was discussed here:

-   week06-treatment-mapping-part1.qmd

-   week06-treatment-mapping-part2.qmd

Total time estimate: Two hours

## Load Packages and Data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor) 
library(arrow) # lets us read/write parquet and feather files
library(sf) # "simple features" for spatial data
library(urbnmapr) # access to US state/country shapefiles for mapping
library(naniar) # tools for exploring and visualizing missing data
library(ggiraph) # makes ggplot charts interactive
options(scipen = 99) # turns off scientific notation

```

```{r}
teds_d <- read_parquet(here("data/teds_d_clean.parquet"))
```

## Question 1

Make an interactive map with `ggiraph` showing the percentage of completed treatments that end with no use at discharge. I am looking for a single interactive map as the final product.

Hint: Break this question down into bite-sized steps, like we did in Part 2. The `freq1_d` column contains information about use at discharge. Check your denominator carefully — should it be *all cases* or *completed cases*?

```{r}
map <- left_join(
  get_urbn_map("states", sf = TRUE),
  teds_d |>
    mutate(across(c(reason, freq1_d), ~ str_squish(str_to_lower(.x)))) |>
    filter(str_detect(reason, "completed")) |>
    group_by(stfips) |>
    summarise(
      pct_no_use = 100 * mean(str_detect(freq1_d, "^(no|none)\\b|no\\s*(substance)?\\s*use|abstin"), na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(state_fips = str_pad(as.character(stfips), 2, pad = "0")),
  by = "state_fips"
)

girafe(ggobj =
  ggplot(map) +
    geom_sf_interactive(aes(geometry = geometry, fill = pct_no_use,
                            tooltip = paste0(state_name, ": ",
                                             ifelse(is.na(pct_no_use),"No data", round(pct_no_use,1)), "%")),
                        color = "white", size = 0.25) +
    scale_fill_viridis_c(na.value = "grey90") +
    labs(title = "Completed Treatments Ending with No Use", fill = "% No Use") +
    theme_minimal()
)

```

## Question 2

Treatment outcomes can vary depending on the type of service patients receive and the length of stay in treatment.

-   Investigate how the percentage of treatments completed and the percentage of treatments that end with no use at discharge differ when you look at service type and length of stay.

-   Create at least three different visualizations to help answer this question.

-   At least one visualization should focus on service type and at least one should focus on length of stay. A third visualization should compare the two.

-   After making your visualizations, write a short summary (3-ish sentences) that explains the main patterns you observe. Please use full, grammatical sentences.

```{r}
service_summary <- teds_d %>%
  group_by(services_d) %>%
  summarise(
    total = n(),
    pct_completed = 100 * mean(reason == "Treatment completed", na.rm = TRUE),
    pct_no_use = 100 * mean(freq1_d == "No use", na.rm = TRUE)
  )

# get completion and no-use percentages by length of stay
los_summary <- teds_d %>%
  group_by(los) %>%
  summarise(
    total = n(),
    pct_completed = 100 * mean(reason == "Treatment completed", na.rm = TRUE),
    pct_no_use = 100 * mean(freq1_d == "No use", na.rm = TRUE)
  )

# combined summary by both
combined_summary <- teds_d %>%
  group_by(services_d, los) %>%
  summarise(
    pct_completed = 100 * mean(reason == "Treatment completed", na.rm = TRUE),
    pct_no_use = 100 * mean(freq1_d == "No use", na.rm = TRUE)
  )

ggplot(service_summary, aes(x = reorder(services_d, pct_completed), y = pct_completed)) +
  geom_col(fill = "#4C9BE8", width = 0.7) +
  coord_flip() +
  labs(
    title = "Percent of Treatments Completed by Service Type",
    x = "Service Type",
    y = "% Completed"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10)
  )

ggplot(combined_summary, aes(x = los, y = pct_completed, fill = services_d)) +
  geom_col(position = "dodge") +
  labs(title = "Completion % by Service Type and Length of Stay",
       x = "Length of Stay", y = "% Completed", fill = "Service Type") +
  theme_minimal()

view(service_summary)
view(los_summary)
view(combined_summary)
```

Summary: People in inpatient and residential programs were more likely to finish their treatment. Outpatient programs had the lowest completion rates, meaning people who go home each day may have a harder time staying in treatment. Overall, programs that give more support seem to help people complete their treatment more often.

## Push to GitHub to Submit!

Assignment is due Tuesday 10/7 at 11:59 pm.
