---
title: "Treatment Mapping"
author: "Amber Camp" 
date: "`r Sys.Date()`"
format: pdf
editor: visual
---

## Load Libraries

If you don't have the below packages, run the relevant lines here:

```{r}
# install.packages("arrow")
# install.packages("sf")
# install.packages("remotes") # used to install packages from github
# remotes::install_github("UrbanInstitute/urbnmapr")
# install.packages("naniar")
# install.packages("ggiraph")
```

Load libraries (all of them).

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor) 
library(arrow) # lets us read/write parquet and feather files
library(sf) # "simple features" for spatial data
library(urbnmapr) # access to US state/country shapefiles for mapping
library(naniar) # tools for exploring and visualizing missing data
library(ggiraph) # makes ggplot charts interactive
options(scipen = 99) # turns off scientific notation
```

## Today's Data

The data we will analyze today is the Treatment Episode Data Set-Discharges (TEDS-D) Dataset from the Substance Abuse and Mental Health Services Administration (SAMHSA). The metadata can be found [here](https://www.samhsa.gov/data/system/files/media-puf-file/TEDS-D-2021-DS0001-info-codebook.pdf)

This is a huge dataset! It's been compressed to a parquet file, which we can read in using `arrow`.

```{r, echo=FALSE}
teds_d <- read_parquet(here("data/teds_d_lecture.parquet"))

teds_d <- teds_d %>%
  clean_names()

head(teds_d)
```

### Columns for today's class

-   Frequency of use at discharge = `freq1_d`

-   State = `stfips` (FIPS = *Federal Information Processing Standard)*

-   Treatment Service = `services_d`

-   Length of Stay = `los`

-   Reason for Discharge = `reason`

### NA Analysis

What data is missing? Examine and visualize below with `naniar`

```{r}
miss_var_summary(teds_d)

gg_miss_var(teds_d)

teds_d %>%
  slice_sample(n = 500) %>% # from dplyr
  vis_miss() # from naniar
```

There actually is more missing data than the above would suggest... How does the documentation label missing data?

```{r}
teds_d %>% 
  tabyl(reason) # from janitor package
```

Replace values as needed.

```{r}

```

Inspect once again.

```{r}

```

### Variable Re-coding

Many of the variables are currently in a format that is not helpful for our analyses.

#### Frequency of Use at Discharge

Start by inspecting this variable. Next, look up the Frequency of Use at Discharge variable in the codebook (`freq1_d`)

```{r}

```

#### Services

Apply the relevant recoding for `services_d`

```{r}

```

#### Reason

Apply the relevant recoding for the Reason for Discharge variable.

```{r}

```

## Quick Plotting Exercise

Create three basic visualizations using `ggplot2` and the variables you recoded above.

```{r}

```

## Do before Wednesday's Class

### Save your new dataframe

You've put a lot of work into cleaning up this dataframe! Once you're done recoding everything, write to a new parquet file so that you can simply read it in on Wednesday!

Have this done before class on Wednesday.

```{r}
write_parquet(teds_d, here("data/teds_d_clean.parquet"))
```
